---
title: "Logistic Regression (and other nonlinear models)"
subtitle: "EDS 222"
author: "Tamma Carleton"
#institute: ""
date: "Fall 2023"
output:
  xaringan::moon_reader:
    css: ['default', 'metropolis', 'metropolis-fonts', 'my-css.css']
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
```{r Setup, include = F}
options(htmltools.dir.version = FALSE)
library(pacman, dplyr)
p_load(leaflet, ggplot2, ggthemes, viridis, dplyr, magrittr, knitr,pagedown,cowplot,latex2exp,parallel,gg3D)

rootdir <- ("~/Dropbox/Teaching/UCSB/EDS_222/EDS222_data")
source(file.path(rootdir,"labs","_common.R"))

# Define pink color
red_pink <- "#e64173"
# Notes directory
dir_slides <- "~/Dropbox/Teaching/UCSB/EDS_222/EDS222_code/EDS-222-stats/slides/06-nonlinearmodels/"
# Knitr options
opts_chunk$set(
  comment = "#>",
  fig.align = "center",
  fig.height = 7,
  fig.width = 10.5,
  # dpi = 300,
  # cache = T,
  warning = F,
  message = F,
  dev = "svg"
)
# A blank theme for ggplot
theme_empty <- theme_bw() + theme(
  line = element_blank(),
  rect = element_blank(),
  strip.text = element_blank(),
  axis.text = element_blank(),
  plot.title = element_blank(),
  axis.title = element_blank(),
  plot.margin = structure(c(0, 0, -1, -1), unit = "lines", valid.unit = 3L, class = "unit"),
  legend.position = "none"
)
theme_simple <- theme_bw() + theme(
  line = element_blank(),
  panel.grid = element_blank(),
  rect = element_blank(),
  strip.text = element_blank(),
  axis.text.x = element_text(size = 14),
  axis.text.y = element_blank(),
  axis.ticks = element_blank(),
  plot.title = element_blank(),
  axis.title = element_blank(),
  # plot.margin = structure(c(0, 0, -1, -1), unit = "lines", valid.unit = 3L, class = "unit"),
  legend.position = "none"
)

theme_axes <- theme_empty + theme(
  axis.title = element_text(size = 18),
  plot.margin = structure(c(0, 0, 0.1, 0), unit = "lines", valid.unit = 3L, class = "unit"),
)

source("../../labs/_common.R")

```

```{css, echo = FALSE}
@media print {
  .has-continuation {
    display: block;
  }
}
```

# Announcements/check-in

- Assignment 03 pass/fail, due **today** (5pm)

--

- Assignment 04 after we cover inference/uncertainty (likely assigned next week)

--

- Final project proposals, due 11/10 (5pm)
  + More details in a few slides


---
# Final project


### Goal:

Apply **some of** the statistical concepts you have learned in this course to **answer an environmental data science question**.<sup>*</sup> 

--

### Two parts:

Deliverable 1: Technical blog post. Some examples:
  + [G-FEED](http://www.g-feed.com/2020/09/indirect-mortality-from-recent.html)
  + [emLab](https://emlab.ucsb.edu/blog/summertime-blues)
  + [MEDS '22, ex. 1](https://cullen-molitor.github.io/posts/2021-12-05-species-density-sst-lagsst/)
  + [MEDS '22, ex. 2](https://jake-eisaguirre.github.io/posts/2021-11-29-mpasandkelp/)
  + [MEDS '22, ex. 3](https://hdolinh.github.io/posts/2021-11-14-stats-final/)

---
# Final project


### Goal:

Apply **some of** the statistical concepts you have learned in this course to **answer an environmental data science question**.<sup>*</sup> 

--

### Two parts:
  
Deliverable 2: Three-minute in-class presentation during final exam slot (4-7pm, 12/12)


.footnote[
[*]: Your project _must_ include concepts from the second half of the course.
] 

---
# Final project

### Proposal: 

Short paragraph (4-5 sentences) describing your proposed project. Motivate the question, describe possible data sources, suggest possible analyses. 

**Email Sandy your proposal** at sandysum@ucsb.edu by 5pm on November 10th. 

---
# Final project

Full guidelines on our [Resources Page](https://tcarleton.github.io/EDS-222-stats/resources.html)

### Some example topics:

- Are political views on climate change associated with recent natural disaster exposure? 

--

- Detecting trends in water quality for indigenous communities in Chile

--

- Spatial patterns of deforestation during COVID-19

--

- Are there gendered health effects of wildfire smoke?

---
name: Overview

# Today

#### More on nonlinear relationships with linear regression models
Log-linear, log-log regressions

--

#### Logistic regression
How do we model binary outcomes? 

---
layout: false
class: clear, middle, inverse
# Nonlinear relationships in linear regression models

---
# Nonlinear transformations

- Our linearity assumption requires that **parameters enter linearly** (_i.e._, the $\beta_k$ multiplied by variables)
- We allow nonlinear relationships between $y$ and the explanatory variables $x$.

**Example: Polynomials**

$$y_i = \beta_0 + \beta_1 x_i + \beta_2 x_i^2 + u_i$$

$$y_i = \beta_0 + \beta_1 x_i + \beta_2 x_i^2 + \beta_3 x_i^3 + u_i$$

$$y_i = \beta_0 + \beta_1 x_i + \beta_2 x_i^2 + \beta_3 x_i^3 + \beta_4 x_i^4 + u_i$$

...
---
# Polynomials

- Recall the relationship between **temperature** and **harmful algal blooms**:

$$ area_i = \beta_0 + \beta_1 temperature_i + \beta_2 temperature_i^2 + u_i$$

```{R, polys, dev = "svg", out.width="70%", echo = F}
# Set seed
set.seed(1235)
# Sample size
n <- 1e3
# Generate data
blooms_df <- tibble(
  temp = runif(n, 0, 3),
  # y = 1 + x + x^2 + x^3 + x^4 + 0.5 * x^5 + rnorm(n, mean = 0, sd = 6)
  area = 0.5*temp + 2*temp^2 + rnorm(n, mean = 0, sd=3)
)
# Plot
ggplot(data = blooms_df, aes(x = temp, y = area)) +
  geom_point(size = 2.5, color = "darkslategrey", alpha = 1) +
  labs(x="temperature (degrees C)", y = "bloom area (hectares)") + 
  theme_bw() + 
  theme(
    line = element_blank(),
  panel.grid = element_blank(),
  rect = element_blank(),
  strip.text = element_blank(),
  axis.text.x = element_blank(),
  axis.text.y = element_blank(),
  axis.title = element_text(size=12,face="bold"),
  plot.title = element_blank(),
  legend.position = "none")
```
---
# Polynomials

Estimating polynomial regressions in `R`:
```{r, echo = T}
blooms_df = blooms_df %>% mutate(temp2 = temp^2)
summary(lm(area~temp+temp2, data=blooms_df))
```
---
# Other nonlinear-in-X regressions

- **Polynomials** and **interactions:** $y_i = \beta_0 + \beta_1 x_{1i} + \beta_2 x_{1i}^2 + \beta_3 x_{2i} + \beta_4 x_{2i}^2 + \beta_5 \left( x_{1i} x_{2i} \right) + u_i$ (more on this today)

- **Exponentials** $\log(y_i) = \beta_0 + \beta_2 e^{x_{2i}} + u_i$ 

- **Logs:** $\log(y_i) = \beta_0 + \beta_1 x_{1i} + u_i$ (Today!)

- **Indicators** and **thresholds:** $y_i = \beta_0 + \beta_1 x_{1i} + \beta_2 \, \mathbb{I}(x_{1i} \geq 100) + u_i$

--

In all cases, the effect of a change in $x$ on $y$ will vary depending on your baseline level of $x$. This is not true with linear relationships! 


---

# Log-linear specification

You will frequently see logged<sup>*</sup>  outcome variables with linear (non-logged) explanatory variables, _e.g._,

$$ \log(\text{area}_i) = \beta_0 + \beta_1 \, \text{temperature}_i + u_i $$

This specification changes our interpretation of the slope coefficients.

--

**Interpretation**

- A one-unit increase in our explanatory variable increases the outcome variable by approximately $\beta_1\times 100$ percent.

- *Example:* If $\beta_1 = 0.03$, an additional degree of warming increases algal bloom area by approximately 3 percent.

.footnote[
[*]: When I say "log", I mean "natural log", i.e. $ln(x) = log_e(x)$.
] 

---
# Review: Percent changes

- What is a percent change again, anyway?

--

- Local gasoline prices were $5/gallon, but last month increased by 12%. How much are they now?

--

$$ 5(1+0.12) = 5\times1.12 = 5.6$$ 
--

Can also write this as $$0.12 = \frac{5.6-5}{5}$$

--

Generally, we have that when $y$ increases by $r$ percent, our new value is $y(1+r)$.

$$ r = \frac{y_2 - y_1}{y_1}$$
---
# Log differences as percent changes?

Near $y=1$, $log(y)$ is approximately slope 1, i.e. $log(y) \approx y-1$


```{R, logs, dev = "svg", out.width = "80%", echo = F}

x = seq(0.1, 4, by=0.01)
lny = log(x)
df = data.frame(x,lny)
# Plot
ggplot(data = df, aes(x = x, y = lny)) +
  geom_line(size = 2.5, color = "darkslategrey", alpha = 1) +
  labs(x="y", y = "log(y), y - 1") + geom_abline(intercept = -1, slope = 1, color = "lightcoral", size=2, linetype="dashed") +
  theme_bw() + 
  theme(
  rect = element_blank(),
  strip.text = element_blank(),  
  axis.text.x = element_text(size=12),
  axis.text.y = element_text(size=12),
  axis.title = element_text(size=12,face="bold"),
  plot.title = element_blank(),
  legend.position = "none")
```

---

# Log differences as percent changes?

Near $y=1$, $log(y)$ is approximately slope 1, i.e. $log(y) \approx y-1$

Therefore, $log(1+r) \approx r$ **when $r$ is small!** (so that you're still close to 1 on the x-axis)

--

This lets us show that:

$$log(y(1+r))  =  log(y) + log(1+r) \approx log(y) + r$$

So when we see $log(y)$ go up by $r$, we can say that represents an $r \times 100$ percent change in $y$!

--

For example: $y$ is increased by 5% means $y$ increases to $y(1.05)$. The log of $y$ changes from $log(y)$ to approximately $log(y) + 0.05$.  Increasing $y$ by 5% is therefore (almost) equivalent to adding 0.05 to $log(y)$.

---
# Log-linear specification

Back to our log-linear model

$$ \log(y_i) = \beta_0 + \beta_1 \, x_i + u $$

A one unit change in $x$ causes a $\beta_1$ unit change in $log(y)$. 

This is equivalent to a $\beta_1$ **percentage change** in $y$.

---
# Log-linear specification

Because the log-linear specification comes with a different interpretation, you need to make sure it fits your data-generating process/model.

Does $x$ change $y$ in levels (_e.g._, a 3-unit increase) or percentages (_e.g._, a 10-percent increase)?

--

_I.e._, you need to be sure an exponential relationship makes sense:

$$ \log(y_i) = \beta_0 + \beta_1 \, x_i + u_i \iff y_i = e^{\beta_0 + \beta_1 x_i + u_i} $$

Note: You are using linear regression to estimate a nonlinear-in-parameters relationship. This is the power of taking logs! 

---
# Log-linear specification

```{r, log linear plot, echo = F, cache = T, dev = "svg", fig.height = 6}
# Set seed
set.seed(1234)
# Sample size
n <- 1e3
# Generate data
ll_df <- tibble(
  x = runif(n, 0, 3),
  y = exp(-100 + 0.75 * x + rnorm(n, sd = 0.5))
)
# Plot
ggplot(data = ll_df, aes(x = x, y = y)) +
geom_hline(yintercept = 0) +
geom_vline(xintercept = 0) +
geom_point(size = 2, color = "darkslategrey", alpha = 0.5) +
geom_smooth(color = red_pink, se = F, size=2) +
xlab("x") +
ylab("y") +
theme_axes
```


---
# Log-log specification

Similarly, log-log models are those where the outcome variable is logged *and* at least one explanatory variable is logged

$$ \log(\text{log}_i) = \beta_0 + \beta_1 \, \log(\text{temperature}_i) + u_i $$

**Interpretation:**

- A one-percent increase in $x$ will lead to a $\beta_1$ percent change in $y$.
- Often interpreted as an "elasticity" in economics.

---
# Log-log specification

```{r, log log plot, echo = F, cache = T, dev = "svg", fig.height = 6}
# Set seed
set.seed(1234)
# Sample size
n <- 1e3
# Generate data
log_df <- tibble(
  x = runif(n, 0, 10),
  y = exp(3 * log(x) + rnorm(n, sd = 0.5))
)
# Plot
ggplot(data = log_df, aes(x = x, y = y)) +
geom_hline(yintercept = 0) +
geom_vline(xintercept = 0) +
geom_point(size = 2, color = "darkslategrey", alpha = 0.5) +
geom_smooth(color = red_pink, se = F, size=2) +
xlab("x") +
ylab("y") +
theme_axes
```
---
# Log-linear with a binary variable

**Note:** If you have a log-linear model with a binary indicator variable, the interpretation for the coefficient on that variable changes.

Consider:

$$\log(y_i) = \beta_0 + \beta_1 x_{1i} + u_i$$

for binary variable $x_1$.

The interpretation of $\beta_1$ is now

- When $x_1$ changes from 0 to 1, $y$ will change by $100 \times \left( e^{\beta_1} -1 \right)$ percent.
- When $x_1$ changes from 1 to 0, $y$ will change by $100 \times \left( e^{-\beta_1} -1 \right)$ percent.


---
# When the approximation fails

The nice interpretation so far relies on the fact that near 1, $log(y) \approx y-1$ 
  - So, for example, $log(y(1+r)) = log(y) + log(1+r) \approx log(y) + r$

--

What if $r$ is large? E.g., $r$=0.8:
  - $log(1*(1.8)) = log(1) + log(1.8) = 0.59 \neq log(1) + 0.8 = 0.8$

--

Exact percentage change (use for large predicted changes):

If $log(y) = \beta_0 + \beta_1 x + \varepsilon$, then the percentage change in $y$ for a one unit change in $x$ is:

$$\text{% change in y} = (e^{\beta_1}-1)\times 100 $$
--

Note that $e^x$ in `R` is `exp(x)`

---
# When the approximation fails

Example: Suppose in $log(y) = \beta_0 + \beta_1 x + \varepsilon$, we estimate that $\hat\beta_1 = 0.6$ 

--

This looks like a 1 unit change in $x$ causes a 60% change in $y$. But the exact percentage change in $y$ is:
  + $(e^{0.6}-1)\times 100 = 0.82 \times 100 \implies 82$ percent change in $y$
  + Note that the imprecise approximation for large changes will always be biased _downwards_

--

Can you just change units of $x$? 
  + Yes, mechanically you can do this and avoid the issues with approximation
  + But think hard about your problem! You probably care about understanding the impacts of a meaningful increase in $x$, not a tiny increase in $x$ 
  
---
layout: false
class: clear, middle, inverse
# Logistic regression
---
# Modeling binary outcomes 

What do you do when your dependent variable takes on just two values? 

```{r, binary, echo = F, cache = T, dev = "svg", out.width = "90%"}
# Set seed
set.seed(1234)
n <- 100
x <- round(runif(100, min = 0, max = 50))
xb <- -9 + 0.4*x 
p <- 1/(1 + exp(-xb))
y <- rbinom(n = 100, size = 1, prob = p)
log_df = tibble(x,y,p)

# Plot
ggplot(data = log_df, aes(x = x, y = y)) +
geom_hline(yintercept = 0) +
geom_vline(xintercept = 0) +
geom_point(size = 3, color = "darkslategrey", alpha = 1) +
xlab("forest cover (%)") +
ylab("species presence") +
  theme_bw() + 
  theme(
  rect = element_blank(),
  strip.text = element_blank(),  
  axis.text.x = element_text(size=12),
  axis.text.y = element_text(size=12),
  axis.title = element_text(size=12,face="bold"),
  plot.title = element_blank(),
  legend.position = "none")
```

---
# Modeling binary outcomes 

What's wrong with running our standard linear regression?

$$\text{species present}_i = \beta_0  + \beta_1 \text{forest cover}_i + \varepsilon_i $$
```{r, lpm, echo = F, cache = T, dev = "svg", out.width = "90%"}
# Plot
ggplot(data = log_df, aes(x = x, y = y)) +
geom_hline(yintercept = 0) +
geom_vline(xintercept = 0) +
geom_point(size = 3, color = "darkslategrey", alpha = 1) +
  geom_smooth(method="lm", se = F, size=2, color = "lightcoral", alpha=1) +
xlab("forest cover (%)") +
ylab("species presence") +
  theme_bw() + 
  theme(
  rect = element_blank(),
  strip.text = element_blank(),  
  axis.text.x = element_text(size=12),
  axis.text.y = element_text(size=12),
  axis.title = element_text(size=12,face="bold"),
  plot.title = element_blank(),
  legend.position = "none")
```

---
# Modeling probabilities

- Our data take on the form $y_i = 1$ or $y_i = 0$

--

- For each individual $i$, there is some probability $p_i$ they have $y_i=1$, so probability $1-p_i$ they have $y_i=0$

--

- We are interested in how a change in variable $x$ changes the probability of $y_i=1$
  + That is, **we model $p_i$** as a function of independent variables
  
--

- Basic idea: we need some transformation of the _probability_ that lets us write:

$$transformation(p_i) = \beta_0 + \beta_1 x_{1i} + \beta_2 x_{2i} + ...$$

---
# Modeling probabilities

Basic idea: we need some transformation of the _probability_ that lets us write:

$$transformation(p_i) = \beta_0 + \beta_1 x_{1i} + \beta_2 x_{2i} + ...$$

- We want this transformation to ensure that:
  + we can input a value between 0 and 1 and return a continuous variable (i.e., we want our outcome variable to be a continuous variable)
  + our predicted probabilities $\hat p_i$ (inverse of the transformation) will fall between 0 and 1

---
# Logistic regression

The **logit function** is the most commonly used nonlinear transformation that ensures predicted probabilities between 0 and 1:

```{r, logit, echo = F, cache = T, dev = "svg", out.width = "90%"}
logitdf <- tibble(
  logitp = seq(-6, 6, 1),
  p = exp(logitp)/(1+exp(logitp))
)

ggplot(data = logitdf, aes(x = logitp, y = p)) +
geom_hline(yintercept = 0) +
geom_vline(xintercept = 0) +
  geom_line(size=2, color = "lightcoral", alpha=1) +
  geom_point(size = 6, color = "darkslategrey", alpha = 1) +
xlab("logit(p)") +
ylab("p") +
  theme_bw() + 
  theme(
  rect = element_blank(),
  strip.text = element_blank(),  
  axis.text.x = element_text(size=12),
  axis.text.y = element_text(size=12),
  axis.title = element_text(size=12,face="bold"),
  plot.title = element_blank(),
  legend.position = "none")
```

---
# Logistic regression

The **logit function** is the most commonly used nonlinear transformation that ensures predicted probabilities between 0 and 1:

$$logit(p) = log\left(\frac{p}{1-p}\right)$$

--

We can then write:

$$log\left(\frac{p_i}{1-p_i}\right) = \beta_0 + \beta_1 x_{1i} + \beta_2 x_{2i} + ...$$

The logit function is also called "log odds" because the "odds ratio" is the probability of success, $p_i$, divided by the probability of failure, $1-p_i$

--

Because of the properties of the logit function (see last graph), this ensures we will generate predicted probabilities $\hat{p}_i$ that fall between 0 and 1.

---
# Logistic regression

How do we estimate this regression?

$$log\left(\frac{p_i}{1-p_i}\right) = \beta_0 + \beta_1 x_{1i} + \beta_2 x_{2i} + ...$$

--

- Can't use linear regression -- we don't have data on $p_i$! We only see $y_i = 1$ or $y_i = 0$
- We use what's called "maximum likelihood estimation" (alternatively, can use gradient descent)
  + Essentially, this asks: what combination of parameters $\beta_0, \beta_1, ...$ maximizes the likelihood that we would observe the data we have? 
  + E.g., if you have high $x_1$ values coinciding with many $y_i =1$ values, likely that $\beta_1$ is high and that $p_i$ is high for observations with large $x_1$

---
# Logistic regression

How do we estimate this regression?

$$log\left(\frac{p_i}{1-p_i}\right) = \beta_0 + \beta_1 x_{1i} + \beta_2 x_{2i} + ...$$

All you really need to know on estimation is...
  + That we use `glm()` instead of `lm()` -- GLM for "generalized linear model" 
  + Interpreting coefficients is a lot more complicated! (next slide)

---
# Interpreting logistic regression output

$$log\left(\frac{p_i}{1-p_i}\right) = \beta_0 + \beta_1 x_{1i} + \beta_2 x_{2i} + ...$$

- $\beta_k$: effect of a 1-unit change in $x_k$ on the log-odds of $y = 1$ 🤔

--

We need to transform our output to get predicted probabilities back! 

$$
\begin{aligned}
\log\left( \frac{p_i}{1-p_i} \right) &= b_0 + b_1 x_{1,i} + \cdots + b_k x_{k,i} \\
\frac{p_i}{1-p_i} &= e^{b_0 + b_1 x_{1,i} + \cdots + b_k x_{k,i}} \\
p_i &= \left( 1 - p_i \right) e^{b_0 + b_1 x_{1,i} + \cdots + b_k x_{k,i}} \\
p_i &= e^{b_0 + b_1 x_{1,i}  + \cdots + b_k x_{k,i}} - p_i \times e^{b_0 + b_1 x_{1,i} + \cdots + b_k x_{k,i}} \\
p_i + p_i \text{ } e^{b_0 + b_1 x_{1,i} + \cdots + b_k x_{k,i}} &= e^{b_0 + b_1 x_{1,i} + \cdots + b_k x_{k,i}} \\
p_i(1 + e^{b_0 + b_1 x_{1,i} + \cdots + b_k x_{k,i}}) &= e^{b_0 + b_1 x_{1,i} + \cdots + b_k x_{k,i}} \\
p_i &= \frac{e^{b_0 + b_1 x_{1,i}  + \cdots + b_k x_{k,i}}}{1 + e^{b_0 + b_1 x_{1,i} + \cdots + b_k x_{k,i}}}
\end{aligned}
$$
---
# Interpreting logistic regression output

This means that if you run a regression with many independent variables, you need to plug your estimated $\hat\beta$'s _and_ the values of all your $x$ variables into this equation to get back a predicted probability for any individual:

$$p_i = \frac{e^{b_0 + b_1 x_{1,i}  + \cdots + b_k x_{k,i}}}{1 + e^{b_0 + b_1 x_{1,i} + \cdots + b_k x_{k,i}}}$$
--

If you want to know the _effect_ of changing just one variable $x_j$ on the probability $p_i$, you need to compute:

$$
\begin{aligned}
p_i(x_j+1) - p_i(x_j) &= \frac{e^{b_0 + \cdots + b_j (x_{j,i}+1)  + \cdots + b_k x_{k,i}}}{1 + e^{b_0 + \cdots + b_j (x_{j,i}+1) + \cdots + b_k x_{k,i}}} - \frac{e^{b_0 + \cdots + b_j x_{j,i}  + \cdots + b_k x_{k,i}}}{1 + e^{b_0 + \cdots + b_j x_{j,i} + \cdots + b_k x_{k,i}}}
\end{aligned}
$$

--

**Note** that this calculation depends on all the other $x$'s! And it will vary with the baseline level of $x_j$ 

---
# Logistic regression: Example

- Bertrand and Mullainathan (2003) study discrimination in hiring decisions
- Authors created many fake resumes, randomly assigning different characteristics (name, sex, race, experience, honors, etc.)

--

- **Outcome variable is binary:** Did the resume get a call back from a (real) potential employer? 
  + Yes: $y_i=1$
  + No: $y_i=0$
  

- Manipulated first names to be those that are commonly associated with White or Black individuals
- Random study design allows estimation of the causal effect of race on callback probability

---
# Logistic regression: Example

```{r resume-names}
resume_names_full <- resume %>% 
  select(firstname, race, gender) %>% 
  distinct(firstname, .keep_all = TRUE) %>% 
  arrange(firstname) %>%
  rownames_to_column() %>% 
  mutate(
    rowname = as.numeric(rowname), 
    column = cut(rowname, breaks = c(0, 12, 24, 36)),
    race = str_to_title(race),
    sex = if_else(gender == "f", "female", "male"),
    column = as.numeric(column)
    ) %>%
  select(-rowname, -gender) %>%
  relocate(column)
resume_names_1 <- resume_names_full %>% 
  filter(column == 1) %>%
  select(-column)
resume_names_2 <- resume_names_full %>% 
  filter(column == 2) %>%
  select(-column)
resume_names_3 <- resume_names_full %>% 
  filter(column == 3) %>%
  select(-column)
resume_names_1 %>%
  bind_cols(resume_names_2) %>%
  bind_cols(resume_names_3) %>%
  kbl(linesep = "", booktabs = TRUE, align = "lllllllll", 
      col.names = c("first_name", "race", "sex",
                    "first_name", "race", "sex",
                    "first_name", "race", "sex"),
      caption = "List of all 36 unique names along with the commonly inferred race and sex associated with these names.") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), 
                latex_options = c("striped", "hold_position"), full_width = FALSE) %>%
  column_spec(4, border_left = T) %>% 
  column_spec(7, border_left = T)
```
---
# Logistic regression: Example

Variables included in the data (all randomly assigned):

```{r resume-data-prep}
resume <- resume %>%
  rename(sex = gender) %>%
  mutate(
    sex = if_else(sex == "m", "man", "woman"),
    sex = fct_relevel(sex, "woman", "man"),
    received_callback = as.factor(received_callback),
    college_degree = as.factor(college_degree),
    honors = as.factor(honors),
    military = as.factor(military),
    has_email_address = as.factor(has_email_address),
    race = if_else(race == "black", "Black", "White")
    ) %>%
  select(received_callback, job_city, college_degree, years_experience,
         honors, military, has_email_address, race, sex)
```

```{r, resumeexample, out.width="40%"}
resume_variables <- tribble(
  ~variable,           ~description,
  "received_callback", "Specifies whether the employer called the applicant following submission of the application for the job.",
  "job_city",          "City where the job was located: Boston or Chicago.",
  "college_degree",    "An indicator for whether the resume listed a college degree.",
  "years_experience",  "Number of years of experience listed on the resume.",
  "honors",            "Indicator for the resume listing some sort of honors, e.g. employee of the month.",
  "military",          "Indicator for if the resume listed any military experience.",
  "has_email_address", "Indicator for if the resume listed an email address for the applicant.",
  "race",              "Race of the applicant, implied by their first name listed on the resume.",
  "sex",               "Sex of the applicant (limited to only and in this study), implied by the first name listed on the resume."
)

resume_variables[1:5,] %>%
  kbl(linesep = "", booktabs = TRUE) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"),
                latex_options = c("striped",  "font_size"=7), full_width = TRUE) %>%
  column_spec(1, monospace = TRUE) %>%
  column_spec(2, width = "30em")
```

---
# Logistic regression: Example

Variables included in the data (all randomly assigned):

```{r, resumeexample2, out.width="40%"}

resume_variables[6:9,] %>%
  kbl(linesep = "", booktabs = TRUE) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"),
                latex_options = c("striped",  "font_size"=7), full_width = TRUE) %>%
  column_spec(1, monospace = TRUE) %>%
  column_spec(2, width = "30em")
```

---
# Logistic Regression: example

- First, we estimate a single predictor: `race`
- `race` indicates whether the applicant is White or not (**Note:** `race` is also binary in this case!)
- We find:

$$\log \left( \frac{\widehat{p}_i}{1-\widehat{p}_i} \right) = -2.67 + 0.44 \times {\texttt{race_White}}$$

a.  If a resume is randomly selected from the study and it has a Black associated name, what is the probability it resulted in a callback?

b.  What would the probability be if the resume name was associated with White individuals?

---
# Logistic regression: Example

$$\log \left( \frac{\widehat{p}_i}{1-\widehat{p}_i} \right) = -2.67 + 0.44 \times {\texttt{race_white}}$$


a.  If a resume is randomly selected from the study and it has a Black associated name, what is the probability it resulted in a callback?

--

**Answer:** If a randomly chosen resume is associated with a Black name, then `race_white` takes the value of 0 and the right side of the model equation equals $-2.67$. Solving for $p_i$ gives $log(\frac{\hat p_i}{1-\hat p_i}) = -2.67 \implies \hat p_i = \frac{e^{-2.67}}{1+e^{-2.67}} = 0.065$.   

---
# Logistic regression: Example

$$\log \left( \frac{\widehat{p}_i}{1-\widehat{p}_i} \right) = -2.67 + 0.44 \times {\texttt{race_white}}$$

b.  What would the probability be if the resume name was associated with White individuals?

**Answer:** If the resume had a name associated with White individuals, then the right side of the model equation is $-2.67+0.44\times 1 = -2.23$. This translates into $\hat p_i = 0.097$.

--

**Conclude:** Being White increases the likelihood of a call back, by 3.2 percentage points.

---
# Logistic regression: Example

**Use the same process** to compute predicted probabilities with multiple independent variables, you just have more calculations!

--

For example, you might estimate:

$$
\begin{aligned}
&log \left(\frac{p}{1 - p}\right) \\
&= - 2.7162 - 0.4364 \times \texttt{job_city}_{\texttt{Chicago}} \\
& \quad \quad + 0.0206 \times \texttt{years_experience} \\
& \quad \quad + 0.7634 \times \texttt{honors} - 0.3443 \times \texttt{military} + 0.2221 \times \texttt{email} \\
& \quad \quad + 0.4429 \times \texttt{race}_{\texttt{White}} - 0.1959 \times \texttt{sex}_{\texttt{man}} 
\end{aligned}
$$

To predict callback probability for a White individual, you also need to know job location, experience, honors, military experience, whether they have an email, race, and sex!

---
# Logistic regression: Example

For example, you might estimate:

$$
\begin{aligned}
&log \left(\frac{p}{1 - p}\right) \\
&= - 2.7162 - 0.4364 \times \texttt{job_city}_{\texttt{Chicago}} \\
& \quad \quad + 0.0206 \times \texttt{years_experience} \\
& \quad \quad + 0.7634 \times \texttt{honors} - 0.3443 \times \texttt{military} + 0.2221 \times \texttt{email} \\
& \quad \quad + 0.4429 \times \texttt{race}_{\texttt{White}} - 0.1959 \times \texttt{sex}_{\texttt{man}} 
\end{aligned}
$$


Note: the effect of race on call back now varies based on all the other covariates! 
  + Try it: Effect of being white for Chicago male with 10 years experience, an email, no honors and no military experience _versus_ a female with the same characteristics? 

---
# Multinomial logistic regression

**What if** your outcome variable is categorical, not binary?

--

For example:
  - Species
  - Socioeconomic status
  - Survey responses
  - ...

--

**Multinomial logistic regression** generalizes the binary logistic regression you've seen here to work for multiple outcome categories
  - Model predicts the probability an individual will fall into each category
  - Beyond the scope of this class, but not a far leap from what you've seen here (lots of online resources -- ask me if you're interested!)

---

class: center, middle


Slides created via the R package [**xaringan**](https://github.com/yihui/xaringan).

Some slide components were borrowed from [Ed Rubin's](https://github.com/edrubin/EC421S20) awesome course materials.

---
exclude: true


```{R, print pdfs, echo = F, eval = F}
pagedown::chrome_print(
  input = "06-nonlinearmodels.html",
  output = "06-nonlinearmodels.pdf",
  timeout = 60
)
```

