---
title: "Logistic Regression (and other nonlinear models)"
subtitle: "EDS 222"
author: "Tamma Carleton"
#institute: ""
date: "Fall 2022"
output:
  xaringan::moon_reader:
    css: ['default', 'metropolis', 'metropolis-fonts', 'my-css.css']
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
```{r Setup, include = F}
options(htmltools.dir.version = FALSE)
library(pacman)
p_load(leaflet, ggplot2, ggthemes, viridis, dplyr, magrittr, knitr,pagedown,cowplot,latex2exp,parallel,gg3D)

rootdir <- ("~/Dropbox/Teaching/UCSB/EDS_222/EDS222_data")
source(file.path(rootdir,"labs","_common.R"))

# Define pink color
red_pink <- "#e64173"
# Notes directory
dir_slides <- "~/Dropbox/Teaching/UCSB/EDS_222/EDS222_code/EDS-222-stats/slides/06-nonlinearmodels/"
# Knitr options
opts_chunk$set(
  comment = "#>",
  fig.align = "center",
  fig.height = 7,
  fig.width = 10.5,
  # dpi = 300,
  # cache = T,
  warning = F,
  message = F,
  dev = "svg"
)
# A blank theme for ggplot
theme_empty <- theme_bw() + theme(
  line = element_blank(),
  rect = element_blank(),
  strip.text = element_blank(),
  axis.text = element_blank(),
  plot.title = element_blank(),
  axis.title = element_blank(),
  plot.margin = structure(c(0, 0, -1, -1), unit = "lines", valid.unit = 3L, class = "unit"),
  legend.position = "none"
)
theme_simple <- theme_bw() + theme(
  line = element_blank(),
  panel.grid = element_blank(),
  rect = element_blank(),
  strip.text = element_blank(),
  axis.text.x = element_text(size = 14),
  axis.text.y = element_blank(),
  axis.ticks = element_blank(),
  plot.title = element_blank(),
  axis.title = element_blank(),
  # plot.margin = structure(c(0, 0, -1, -1), unit = "lines", valid.unit = 3L, class = "unit"),
  legend.position = "none"
)

theme_axes <- theme_empty + theme(
  axis.title = element_text(size = 18),
  plot.margin = structure(c(0, 0, 0.1, 0), unit = "lines", valid.unit = 3L, class = "unit"),
)
```

```{css, echo = FALSE}
@media print {
  .has-continuation {
    display: block;
  }
}
```

# Announcements/check-in

- Assignment 03 pass/fail, due **today** (9am)

--

- Assignment 04 after we cover inference/uncertainty (likely assigned next week)

--

- Final project proposals, due 11/10 (9am)
  + More details in a few slides


---
# Final project


### Goal:

Apply **some of** the statistical concepts you have learned in this course to **answer an environmental data science question**.<sup>*</sup> 

--

### Two parts:

1. Technical blog post. Some examples:
  + [G-FEED](http://www.g-feed.com/2020/09/indirect-mortality-from-recent.html)
  + [emLab](https://emlab.ucsb.edu/blog/summertime-blues)
  + [MEDS '22, ex. 1](https://cullen-molitor.github.io/posts/2021-12-05-species-density-sst-lagsst/)
  + [MEDS '22, ex. 2](https://jake-eisaguirre.github.io/posts/2021-11-29-mpasandkelp/)
  + [MEDS '22, ex. 3](https://hdolinh.github.io/posts/2021-11-14-stats-final/)
  
2. Three-minute in-class presentation during final exam slot (8-11am, 12/6)


.footnote[
[*]: Your project _must_ include concepts from the second half of the course.
] 
---
# Final project

### Proposal: 

Short paragraph (4-5 sentences) describing your proposed project. Motivate the question, describe possible data sources, suggest possible analyses. 

**Email Sandy your proposal** at sandysum@ucsb.edu by 9am on November 10th. 

---
# Final project

Full guidelines on our [Resources Page](https://tcarleton.github.io/EDS-222-stats/resources.html)

### Some example topics:

- Are political views on climate change associated with recent natural disaster exposure? 

--

- Detecting trends in air quality for disadvantaged groups across California

--

- Spatial patterns of deforestation during COVID-19

--

- Are there gendered health effects of wildfire smoke?

---
name: Overview

# Today

#### More on nonlinear relationships with linear regression models
Log-linear, log-log regressions

--

#### Logistic regression
How do we model binary outcomes? + note on generalization to discrete categories

---
layout: false
class: clear, middle, inverse
# Nonlinear relationships in linear regression models

---
# Nonlinear transformations

- Our linearity assumption requires that **parameters enter linearly** (_i.e._, the $\beta_k$ multiplied by variables)
- We allow nonlinear relationships between $y$ and the explanatory variables $x$.

**Example: Polynomials**

$$y_i = \beta_0 + \beta_1 x_i + \beta_2 x_i^2 + u_i$$

$$y_i = \beta_0 + \beta_1 x_i + \beta_2 x_i^2 + \beta_3 x_i^3 + u_i$$

$$y_i = \beta_0 + \beta_1 x_i + \beta_2 x_i^2 + \beta_3 x_i^3 + \beta_4 x_i^4 + u_i$$

...
---
# Polynomials

- Recall the relationship between **temperature** and **harmful algal blooms**:

$$ area_i = \beta_0 + \beta_1 temperature_i + \beta_2 temperature_i^2 + u_i$$

```{R, polys, dev = "svg", fig.height = 6, echo = F}
# Set seed
set.seed(1235)
# Sample size
n <- 1e3
# Generate data
blooms_df <- tibble(
  temp = runif(n, 0, 3),
  # y = 1 + x + x^2 + x^3 + x^4 + 0.5 * x^5 + rnorm(n, mean = 0, sd = 6)
  area = 0.5*temp + 2*temp^2 + rnorm(n, mean = 0, sd=3)
)
# Plot
ggplot(data = blooms_df, aes(x = temp, y = area)) +
  geom_point(size = 2.5, color = "darkslategrey", alpha = 1) +
  labs(x="temperature (degrees C)", y = "bloom area (hectares)") + 
  theme_bw() + 
  theme(
    line = element_blank(),
  panel.grid = element_blank(),
  rect = element_blank(),
  strip.text = element_blank(),
  axis.text.x = element_blank(),
  axis.text.y = element_blank(),
  axis.title = element_text(size=18,face="bold"),
  plot.title = element_blank(),
  legend.position = "none")
```
---
# Polynomials

Estimating polynomial regressions in `R`:
```{r, echo = T}
blooms_df = blooms_df %>% mutate(temp2 = temp^2)
summary(lm(area~temp+temp2, data=blooms_df))
```
---
# Other nonlinear-in-X regressions

- **Polynomials** and **interactions:** $y_i = \beta_0 + \beta_1 x_{1i} + \beta_2 x_{1i}^2 + \beta_3 x_{2i} + \beta_4 x_{2i}^2 + \beta_5 \left( x_{1i} x_{2i} \right) + u_i$ (more on this today)

- **Exponentials** $\log(y_i) = \beta_0 + \beta_2 e^{x_{2i}} + u_i$ 

- **Logs:** $\log(y_i) = \beta_0 + \beta_1 x_{1i} + u_i$ (Today!)

- **Indicators** and **thresholds:** $y_i = \beta_0 + \beta_1 x_{1i} + \beta_2 \, \mathbb{I}(x_{1i} \geq 100) + u_i$

--

In all cases, the effect of a change in $x$ on $y$ will vary depending on your baseline level of $x$. This is not true with linear relationships! 


---

# Log-linear specification

You will frequently see logged<sup>*</sup>  outcome variables with linear (non-logged) explanatory variables, _e.g._,

$$ \log(\text{Pay}_i) = \beta_0 + \beta_1 \, \text{School}_i + u_i $$

This specification changes our interpretation of the slope coefficients.

--

**Interpretation**

- A one-unit increase in our explanatory variable increases the outcome variable by approximately $\beta_1\times 100$ percent.

- *Example:* If $\beta_1 = 0.03$, an additional year of schooling increases pay by approximately 3 percent.

.footnote[
[*]: When I say "log", I mean "natural log", i.e. $ln(x) = log_e(x)$.
] 

---
# Review: Percent changes

- What is a percent change again, anyway?

--

- Local gasoline prices were $5/gallon, but last month increased by 12%. How much are they now?

--

$$ 5(1+0.12) = 5\times1.12 = 5.6$$ 
--

Can also write this as $$0.12 = \frac{5.6-5}{5}$$

--

Generally, we have that when $y$ increases by $r$ percent, our new value is $y(1+r)$.

$$ r = \frac{y_2 - y_1}{y_1}$$
---
# Log differences as percent changes?

Near $y=1$, $log(y)$ is approximately slope 1, i.e. $log(y) \approx y-1$


```{R, logs, dev = "svg", fig.height = .5, echo = F}

x = seq(0.1, 4, by=0.01)
lny = log(x)
df = data.frame(x,lny)
# Plot
ggplot(data = df, aes(x = x, y = lny)) +
  geom_line(size = 2.5, color = "darkslategrey", alpha = 1) +
  labs(x="y", y = "y") + geom_abline(intercept = -1, slope = 1, color = "lightcoral", size=2, linetype="dashed") +
  theme_bw() + 
  theme(
  rect = element_blank(),
  strip.text = element_blank(),  
  axis.text.x = element_text(size=18),
  axis.text.y = element_text(size=18),
  axis.title = element_text(size=18,face="bold"),
  plot.title = element_blank(),
  legend.position = "none")
```

---

# Log differences as percent changes?

Near $y=1$, $log(y)$ is approximately slope 1, i.e. $log(y) \approx y-1$

Therefore, $log(1+r) \approx r$ **when $r$ is small!** (so that you're still close to 1 on the x-axis)

--

This lets us show that:

$$log(y(1+r))  =  log(y) + log(1+r) \approx log(y) + r$$

So when we see $log(y)$ go up by $r$, we can say that represents an $r \times 100$ percent change in $y$!

--

For example: $y$ is increased by 5% means $y$ increases to $y(1.05)$. The log of $y$ changes from $log(y)$ to approximately $log(y) + 0.05$.  Increasing $y$ by 5% is therefore (almost) equivalent to adding 0.05 to $log(y)$.

---
# Log-linear specification

Back to our log-linear model

$$ \log(y_i) = \beta_0 + \beta_1 \, x_i + u $$

A one unit change in $x$ causes a $\beta_1$ unit change in $log(y)$. 

This is equivalent to a $\beta_1$ **percentage change** in $y$.

---
# Log-linear specification

Because the log-linear specification comes with a different interpretation, you need to make sure it fits your data-generating process/model.

Does $x$ change $y$ in levels (_e.g._, a 3-unit increase) or percentages (_e.g._, a 10-percent increase)?

--

_I.e._, you need to be sure an exponential relationship makes sense:

$$ \log(y_i) = \beta_0 + \beta_1 \, x_i + u_i \iff y_i = e^{\beta_0 + \beta_1 x_i + u_i} $$

--

Note: You are using linear regression to estimate a nonlinear-in-parameters relationship. This is the power of taking logs! 

---
# Log-linear specification

```{r, log linear plot, echo = F, cache = T, dev = "svg", fig.height = 6}
# Set seed
set.seed(1234)
# Sample size
n <- 1e3
# Generate data
ll_df <- tibble(
  x = runif(n, 0, 3),
  y = exp(-100 + 0.75 * x + rnorm(n, sd = 0.5))
)
# Plot
ggplot(data = ll_df, aes(x = x, y = y)) +
geom_hline(yintercept = 0) +
geom_vline(xintercept = 0) +
geom_point(size = 3, color = "darkslategrey", alpha = 0.5) +
geom_smooth(color = red_pink, se = F) +
xlab("x") +
ylab("y") +
theme_axes
```


---
# Log-log specification

Similarly, log-log models are those where the outcome variable is logged *and* at least one explanatory variable is logged

$$ \log(\text{Pay}_i) = \beta_0 + \beta_1 \, \log(\text{School}_i) + u_i $$

**Interpretation:**

- A one-percent increase in $x$ will lead to a $\beta_1$ percent change in $y$.
- Often interpreted as an "elasticity" in economics.

---
# Log-linear with a binary variable

**Note:** If you have a log-linear model with a binary indicator variable, the interpretation for the coefficient on that variable changes.

Consider

$$ \log(y_i) = \beta_0 + \beta_1 x_{1i} + u_i $$

for binary variable $x_1$.

The interpretation of $\beta_1$ is now

- When $x_1$ changes from 0 to 1, $y$ will change by $100 \times \left( e^{\beta_1} -1 \right)$ percent.
- When $x_1$ changes from 1 to 0, $y$ will change by $100 \times \left( e^{-\beta_1} -1 \right)$ percent.

---
# Log-log specification

```{r, log log plot, echo = F, cache = T, dev = "svg", fig.height = 6}
# Set seed
set.seed(1234)
# Sample size
n <- 1e3
# Generate data
log_df <- tibble(
  x = runif(n, 0, 10),
  y = exp(3 * log(x) + rnorm(n, sd = 0.5))
)
# Plot
ggplot(data = log_df, aes(x = x, y = y)) +
geom_hline(yintercept = 0) +
geom_vline(xintercept = 0) +
geom_point(size = 3, color = "darkslategrey", alpha = 0.5) +
geom_smooth(color = red_pink, se = F) +
xlab("x") +
ylab("y") +
theme_axes
```
---
layout: false
class: clear, middle, inverse
# Logistic regression
---

class: center, middle


Slides created via the R package [**xaringan**](https://github.com/yihui/xaringan).

Some slide components were borrowed from [Ed Rubin's](https://github.com/edrubin/EC421S20) awesome course materials.

---
exclude: true


```{R, print pdfs, echo = F, eval = F}
pagedown::chrome_print(
  input = "06-inference.html",
  output = "06-inference.pdf",
  timeout = 60
)
```

